---
title: "CMI-PB Team's Data Processing Approach for Second Challenge Dataset"
#date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  BiocStyle::html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    code_folding: show
    number_sections: false
date: "2023-08-21"
---

The training dataset for the second challenge comprises two multi-omics datasets (designated as 2020 and 2021) that require processing and normalization to generate computable matrices suitable for subsequent model development. While the procedures for data processing and normalization are inherently user-specific, the CMI-PB team has devised a data processing method, drawing inspiration from the approach used in the internal 1st CMI-PB challenge. The codebase is also available on GitHub. If you have specific questions, please contact us via Solutions center.

## Download and read 2nd challenge data from CMI-PB website

The data files for the 2nd CMI-PB challenge can be accessed at [<https://www.cmi-pb.org/downloads/cmipb_challenge_datasets/current/2nd_challenge/raw_datasets/>]. They are available for direct file download or as R data objects. For our notebook, we chose to download the data as R data objects. These objects contain both demographical metadata of subjects and experimental data, including:

-   **Plasma antibody titers**: Plasma antibodies against Tdap were measured at all time points using Luminex assay,
-   **Plasma cytokine concentrations**: Plasma cytokine concentration analysis using OLINK assay,
-   **PBMC gene expression:** Gene expression analysis (RNAseq) of bulk peripheral blood mononuclear cells (PBMCs),
-   **PBMC cell frequency:** Cell frequency analysis of PBMC subsets were performed before and after booster vaccination until day 14.

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
source("codebase.R")

master_database_data <- readRDS("master_harmonized_training_data.RDS")

subject_specimen <- master_database_data$subject_specimen  %>% 
  mutate(timepoint = planned_day_relative_to_boost)

training_subject_specimen <- master_database_data$subject_specimen

batch.factors = c("timepoint","infancy_vac","biological_sex","dataset")
data_obj = master_database_data

```

## Antibody titers

```{r}
abtiter_wide_before <- data_obj$plasma_antibody_levels$long %>%
  dplyr::select(isotype_antigen, specimen_id, MFI) %>%
  pivot_wider(names_from = "isotype_antigen", values_from = MFI) %>%
  column_to_rownames("specimen_id")%>%
  t() 

pvca_analysis(abtiter_wide_before, data_obj$subject_specimen, batch.factors, plot_title = "Antibody titer:  Raw data")
# pvca_analysis(abtiter_wide_before, data_obj$subject_specimen, batch.factors, plot_title = "Antibody:\nRaw data")
# anti_p1 <- p1

## Apply data normalization and batch correction
abtiter_data_processed = processAbtiter(data_obj, BatchCorrection = TRUE)

pvca_analysis(abtiter_data_processed$normalized_data, data_obj$subject_specimen, batch.factors, plot_title = "Antibody titer: Normalization")
pvca_analysis(abtiter_data_processed$batchCorrected_data, data_obj$subject_specimen, batch.factors, plot_title = "Antibody titer:  Normalization and batch correction")
# pvca_analysis(abtiter_data_processed$normalized_data, data_obj$subject_specimen, batch.factors, plot_title = "Antibody:\nHarmonized")
# anti_p2 <- p1


```

## Cell frequency Analysis

```{r}
## Before  normalization
cell_wide_before <- data_obj$pbmc_cell_frequency$wide %>%
  column_to_rownames("specimen_id")%>%
  t() 

pvca_analysis(cell_wide_before, data_obj$subject_specimen, batch.factors, plot_title = "Cell frequency:  Raw data")
# pvca_analysis(cell_wide_before, data_obj$subject_specimen, batch.factors, plot_title = "Cell frequency:\nRaw data")
# cell_p1 <- p1

## Apply data normalization and batch correction
cytof_data_processed = processCellFreq(data_obj, BatchCorrection = TRUE)

pvca_analysis(cytof_data_processed$normalized_data, data_obj$subject_specimen, batch.factors, plot_title = "Cell frequency: Normalization")
pvca_analysis(cytof_data_processed$batchCorrected_data, data_obj$subject_specimen, batch.factors, plot_title = "Cell frequency:  Normalization and batch correction")
# pvca_analysis(cytof_data_processed$batchCorrected_data, data_obj$subject_specimen, batch.factors, plot_title = "Cell frequency:\nHarmonized")
# cell_p2 <- p1

```

## Olink Analysis

```{r}
## Before batch correction
olink_wide_before <- data_obj$plasma_cytokine_concentrations$wide  %>%
  column_to_rownames("specimen_id")%>%
  t() 

pvca_analysis(olink_wide_before, data_obj$subject_specimen, batch.factors, plot_title = "Cytokine concetrations:  Raw data")
# pvca_analysis(olink_wide_before, data_obj$subject_specimen, batch.factors, plot_title = "Cytokine:\nRaw data")
# olink_p1 <- p1

## Apply data normalization and batch correction
olink_data_processed = processOlink(data_obj, BatchCorrection = TRUE)

pvca_analysis(cytof_data_processed$normalized_data, data_obj$subject_specimen, batch.factors, plot_title = "Cytokine concetrations: Normalization")

pvca_analysis(cytof_data_processed$batchCorrected_data, data_obj$subject_specimen, batch.factors, plot_title = "Cytokine concetrations: Normalization and batch correction")
# pvca_analysis(olink_data_processed$batchCorrected_data, data_obj$subject_specimen, batch.factors, plot_title = "Cytokine:\nHarmonized")
# olink_p2 <- p1
```

## Gene expression data Analysis

```{r}
rnaseq_countData <- data_obj$pbmc_gene_expression_wide$wide %>%
  column_to_rownames("specimen_id") %>%
  t()  %>%
  as.data.frame()  

colnames(rnaseq_countData) = as.integer(colnames(rnaseq_countData))

rnaseq_metaData <- data_obj$subject_specimen %>%
  filter(specimen_id %in% colnames(rnaseq_countData)) %>%
  mutate(specimen_id1 = specimen_id) %>%
  column_to_rownames("specimen_id1") 

#' Identify genes with tpm >=1 that are present in at least 80% of either the aP(wP) cohort and absent in at least 80% of wP(aP) cohort.
threshold_proportion_greater_than_1 = 0.8

tpm_sum_infancy_subgroup <- rnaseq_countData %>%
  rownames_to_column("versioned_ensembl_gene_id") %>%
  pivot_longer(!versioned_ensembl_gene_id, values_to = "tpm", names_to = "specimen_id") %>%
  mutate(specimen_id = as.integer(specimen_id)) %>%
  left_join(subject_specimen) %>%
  group_by(dataset, versioned_ensembl_gene_id, infancy_vac) %>%
  group_by(versioned_ensembl_gene_id, infancy_vac) %>%
  summarise(proportion_greater_than_1 = mean(tpm >= 1)) %>%
  pivot_wider(names_from = infancy_vac, values_from = proportion_greater_than_1)  %>%
  mutate(gene_meets_criterion_aP = aP >= threshold_proportion_greater_than_1 & wP <= (1 - threshold_proportion_greater_than_1),
         gene_meets_criterion_wP = wP >= threshold_proportion_greater_than_1 & aP <= (1 - threshold_proportion_greater_than_1)
         )  %>%
  filter((gene_meets_criterion_aP == TRUE & gene_meets_criterion_wP == FALSE) || (gene_meets_criterion_aP == FALSE & gene_meets_criterion_wP == TRUE))
  
#' Create a shortlist of genes (tpm >= 1) in at least 30% of the specimens.

tpm_shortlist <- rnaseq_countData %>%
  rownames_to_column("versioned_ensembl_gene_id") %>%
  # filter(versioned_ensembl_gene_id %in% gene_90_38_shortlist$versioned_ensembl_gene_id) %>%
  pivot_longer(!versioned_ensembl_gene_id, values_to = "tpm", names_to = "specimen_id") %>%
  mutate(specimen_id = as.integer(specimen_id)) %>%
  left_join(subject_specimen) %>%
  group_by(versioned_ensembl_gene_id) %>%
  #group_by(versioned_ensembl_gene_id, infancy_vac) %>%
  summarise(proportion = mean(tpm >= 1))  %>%
  filter(proportion >= 0.3)


## Before batch correction
rnaseq_countData_v2 <- rnaseq_countData %>%
  rownames_to_column("versioned_ensembl_gene_id") %>%
  # filter(versioned_ensembl_gene_id %in% gene_90_38_shortlist$versioned_ensembl_gene_id) %>%
  filter(!versioned_ensembl_gene_id %in% tpm_sum_infancy_subgroup$versioned_ensembl_gene_id) %>%
  filter(versioned_ensembl_gene_id %in% tpm_shortlist$versioned_ensembl_gene_id) %>%
  column_to_rownames("versioned_ensembl_gene_id")

mad_2020 <- mad_calculations(rnaseq_countData_v2, data_obj$subject_specimen, c("2020_dataset"))
mad_2021 <- mad_calculations(rnaseq_countData_v2, data_obj$subject_specimen, c("2021_dataset"))
mad_2022 <- mad_calculations(rnaseq_countData_v2, data_obj$subject_specimen, c("2022_dataset"))

intersection1 <- intersect(mad_2020$gene_id, mad_2021$gene_id)
intersection2 <- intersect(mad_2020$gene_id, mad_2022$gene_id)
mad_shotlisted_genes <- intersect(intersection1, intersection2)

rnaseq_countData_v3 <- rnaseq_countData_v2 %>%
  rownames_to_column("versioned_ensembl_gene_id") %>%
  filter(versioned_ensembl_gene_id %in% mad_shotlisted_genes) %>%
  column_to_rownames("versioned_ensembl_gene_id")
  
pvca_analysis_rnaseq(rnaseq_countData_v3, data_obj$subject_specimen, batch.factors, plot_title = "RNASeq: Raw data")
# pvca_analysis_rnaseq(rnaseq_countData_v3, data_obj$subject_specimen, batch.factors, plot_title = "RNASeq:\nRaw data")
# rna_p1 <- p1

batch_lebels = as.data.frame(colnames(rnaseq_countData_v3)) %>%
      rename(specimen_id = starts_with("colnames")) %>%
      mutate(specimen_id = as.integer(specimen_id)) %>%
      left_join(rnaseq_metaData) %>%
      dplyr::select(dataset)
    
rnaseq_batchCorrected = sva::ComBat_seq(as.matrix(rnaseq_countData_v3), batch = batch_lebels$dataset)
    
pvca_analysis_rnaseq(rnaseq_batchCorrected, data_obj$subject_specimen, batch.factors, plot_title = "RNASeq: Batch correction")
# pvca_analysis_rnaseq(rnaseq_batchCorrected, data_obj$subject_specimen, batch.factors, plot_title = "RNASeq:\nHarmonized")
# rna_p2 <- p1


rnaseq_normalised_data = list(
      
      metadata = rnaseq_metaData,
      raw_data = rnaseq_countData_v3,
      batchCorrected_data = rnaseq_batchCorrected
    )
```

## Save normalized data

```{r}
master_normalized_data <- list(

  subject_specimen = training_subject_specimen,
  abtiter_wide = abtiter_data_processed,
  plasma_cytokine_concentrations = olink_data_processed,
  pbmc_cell_frequency = cytof_data_processed,
  pbmc_gene_expression = rnaseq_normalised_data

)

sapply(master_normalized_data, dim)

```

## Save data as RDS File

```{r}
# saveRDS(master_normalized_data, file = "master_processed_training_data.RDS")
# sessioninfo::session_info()
```

```{r}
# library(patchwork)
# 
# anti_p1 + cell_p1 + olink_p1 + rna_p1 + anti_p2 + cell_p2 + olink_p2 + rna_p2 +
#   plot_layout(nrow = 2, guides = "collect") & theme(legend.position = "bottom", 
#                                                     legend.title = element_text(size = 24),
#                                                     legend.text = element_text(size=24)) & 
#   plot_annotation(title = '',
#                     #caption = 'made with patchwork',
#                     theme = theme(plot.title = element_text(size = 16)))

```